systemfit: A Package to Estimate\\ Simultaneous Equation Systems in \proglang{R}

### 2006-02-15 paper/trunk/systemfit_estimation.tex
small last changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@579 591db2bd-e0a0-4d22-8c19-2192dbb7da31
just changed some line breaks (deleted new ones and restored deleted ones) to make it easier to find the changes that Jeff committed in the last revision(s)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@582 591db2bd-e0a0-4d22-8c19-2192dbb7da31
changed a few line breaks that each sentence starts in a new line now
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@591 591db2bd-e0a0-4d22-8c19-2192dbb7da31
several minor revisions in sections Introduct

The Ordinary Least Squares (OLS) estimator of the system 
is obtained by
\begin{equation}
   \widehat{\beta} = \left( X'X \right)^{-1} X'y
\end{equation}

These estimates are efficient only if the disturbance terms are not 
contemporaneously correlated, which means 
\begin{equation}
   \sigma_{ij} = 0 \; \forall \; i \neq j
\end{equation}

If the whole system is treated as one single equation, 
the variance-covariance matrix of the estimated parameters is
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \sigma^2 \left( X'X \right)^{-1}
\end{equation}
with $\sigma^2 = E \left( u' u \right)$.
This assumes that the disturbances of all equations have the
same variance.

@@

The Weighted Least Squares (WLS) estimator of the system 
is obtained by
\begin{equation}
   \widehat{\beta} = \left( X' \Omega^{-1} X \right)^{-1} X' \Omega^{-1} y
\end{equation}
with $\Omega = \Sigma \otimes I$, 
$\sigma_{ij} = 0 \; \forall \; i \neq j$ and
$\sigma_{ii} = E \left( u_i' u_i \right)$.

Like the OLS estimates these estimates are only efficient 
if the disturbance terms are not contemporaneously correlated.

The variance-covariance matrix of the estimated parameters is
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \left( X' \Omega^{-1} X \right)^{-1}
\end{equation}

If no cross-equation parameter restrictions are imposed, 
the parameter estimates are equal to the OLS estimates.

@@

When the disturbances are contemporaneously correlated, a Generalized 
Least Squares (GLS) estimation leads to efficient parameter estimates.
In this case the GLS is generally called °Seemingly Unrelated Regression°
(SUR) (see \citealp{zellner62}). 
Note, while an unbiased OLS or WLS estimation requires only that 
the regressors and the disturbance terms of each single 
equation are uncorrelated,
a consistent SUR estimation requires that all disturbance terms and all 
regressors are uncorrelated:
\begin{equation}
   E \left( u | X \right) = 0
\end{equation}

The SUR estimator can be obtained by:%
\footnote{To calculate $\Omega^{-1}$ it is not necessary to invert 
the (huge) $\Omega$ matrix, 
because it is sufficient to invert the (small) $\Sigma$ matrix and
calculate it by $\Omega^{-1} = \Sigma^{-1} \otimes I$.}

\begin{equation}
   \widehat{\beta} = \left( X' \Omega^{-1} X \right)^{-1} X' \Omega^{-1} y
\end{equation}
with $\Omega = \Sigma \otimes I$ and
$\sigma_{ij} = E \left( u_i' u_j \right)$.

And the variance-covariance matrix of the estimated parameters is
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \left( X' \Omega^{-1} X \right)^{-1}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Two-stage least squares (2SLS)}

If the regressors of one or more equations are correlated 
with the disturbances ($E \left( u_i | X_i \right) \neq 0$), 
the estimated coefficients are biased.
This can be circumvented by an instrumental variable (IV) estimation.
The instrumental variables for each equation $H_i$ 
can be either different or identical for all equations.
The instrumental variables of each equation may not be correlated with 
the disturbance terms of the corresponding equation 
($E \left( u_i | H_i \right) = 0$).

At the first stage new ('fitted') regressors are obtained by
\begin{equation}
   \widehat{X_i} = H_i \left( H_i' H_i \right)^{-1} H_i' X
\end{equation}

At the second stage the unbiased two-stage least squares estimates
of $\beta$ are obtained by:
\begin{equation}
   \widehat{\beta} = \left( \widehat{X}' \widehat{X} \right)^{-1} 
   \widehat{X}' y 
   \label{eq:beta2sls}
\end{equation}

If the whole system is treated as one single equation, 
the variance-covariance matrix of the estimated parameters is
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \sigma^2 \left( \widehat{X}'
   \widehat{X} \right)^{-1}
\end{equation}
with $\sigma^2 = E \left( u' u \right)$.

If the disturbance terms of the individual equations 
are allowed to have different variances, 
the variance-covariance matrix of the estimated parameters is
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \left( \widehat{X}' \Omega^{-1} 
   \widehat{X} \right)^{-1}
\end{equation}
with $\Omega = \Sigma \otimes I$, 
$\sigma_{ij} = 0 \; \forall \; i \neq j$ and
$\sigma_{ii} = E \left( u_i' u_i \right)$.

@@

The Weighted Two-Stage Least Squares (W2SLS) estimator of the system 
is obtained by
\begin{equation}
   \widehat{\beta} = \left( \widehat{X}' \Omega^{-1} \widehat{X} 
   \right)^{-1} \widehat{X}' \Omega^{-1} y
\end{equation}
with $\Omega = \Sigma \otimes I$, 
$\sigma_{ij} = 0 \; \forall \; i \neq j$ and
$\sigma_{ii} = E \left( u_i' u_i \right)$.

The variance-covariance matrix of the estimated parameters is
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \left( \widehat{X}' \Omega^{-1} 
   \widehat{X} \right)^{-1}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Three-stage least squares (3SLS)}

If the regressors are correlated with the disturbances 
($E \left( u | X \right) \neq 0$) and 
the disturbances are contemporaneously correlated, 
an Instrumental Variable (IV) Generalized Least Squares (GLS) 
estimation leads to consistent and efficient estimates. This estimation 
procedure is generally called °Three-stage Least Squares° (3SLS).

The standard 3SLS estimator can be obtained by:
\begin{equation}
   \widehat{\beta} = \left( \widehat{X}' \Omega^{-1} \widehat{X} 
   \right)^{-1} \widehat{X}' \Omega^{-1} y
\end{equation}
with $\Omega = \Sigma \otimes I$ and
$\sigma_{ij} = E \left( u_i' u_j \right)$.

Its variance-covariance matrix is:
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \left( \widehat{X}' \Omega^{-1} 
   \widehat{X} \right)^{-1}
   \label{eq:cov3sls}
\end{equation}

While an unbiased 2SLS or W2SLS estimation requires only that 
the instrumental variables and the disturbance terms of each single 
equation are uncorrelated,
\cite{schmidt90} points out that this estimator is only consistent 
if all disturbance terms and all instrumental variables are uncorrelated:
\begin{equation}
   E \left( u | H \right) = 0
\end{equation}
with
\begin{equation}
   H =
   \left[ \begin{array}{cccc}
      H_1 & 0 & \cdots & 0\\
      0 & H_2 & \cdots & 0\\
      \vdots & \vdots & \ddots & \vdots\\
      0 & 0 & \cdots & H_G
   \end{array}\right] \cdot 
\end{equation}

Since there might be occasions where this cannot be avoided, 
\cite{schmidt90} analyses other approaches to obtain 3SLS estimators:

One of these approaches is based on instrumental variable estimation
(IV-3SLS):
\begin{equation}
   \widehat{\beta} = \left( \widehat{X}' \Omega^{-1} X 
   \right)^{-1} \widehat{X}' \Omega^{-1} y
\end{equation}

The variance-covariance matrix of this IV-3SLS estimator is:
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \left( \widehat{X}' \Omega^{-1} 
   X \right)^{-1}
\end{equation}

Another approach is based on the Generalized Method of Moments (GMM) 
estimator (GMM-3SLS):
\begin{equation}
   \widehat{\beta} = \left( X' H \left( H' \Omega H \right)^{-1}
   H' X \right)^{-1} X' H \left( H' \Omega H \right)^{-1} H' y
   \label{eq:beta3slsGMM}
\end{equation}

The variance-covariance matrix of the GMM-3SLS estimator is:
\begin{equation}
   Var \left[ \widehat{\beta} \right] = 
   \left( X' H \left( H' \Omega H \right)^{-1} H' X \right)^{-1}
\end{equation}

A fourth approach developed by \cite{schmidt90} himself is:
\begin{equation}
   \widehat{\beta} = \left( \widehat{X}' \Omega^{-1} \widehat{X} 
   \right)^{-1} \widehat{X}' \Omega^{-1} 
   H \left( H' H \right)^{-1} H' y 
   \label{eq:beta3slsSchmidt}
\end{equation}

The variance-covariance matrix of this estimator is:
\begin{equation}
   Var \left[ \widehat{\beta} \right] = 
   \left( \widehat{X}' \Omega^{-1}  \widehat{X} \right)^{-1} 
   \widehat{X}' \Omega^{-1} H \left( H' H \right)^{-1} H' \Omega 
   H \left( H' H \right)^{-1} H' \Omega^{-1} \widehat{X}
   \left( \widehat{X}' \Omega^{-1}  \widehat{X} \right)^{-1}
\end{equation}

The econometrics software EViews uses following approach:
\begin{equation}
   \widehat{\beta} = \widehat{\beta}_{2SLS} + 
   \left( \widehat{X}' \Omega^{-1} \widehat{X} \right)^{-1} 
   \widehat{X}' \Omega^{-1} \left( y - X \widehat{\beta}_{2SLS} \right)
\end{equation}
with $\widehat{\beta}_{2SLS}$ is the two-stage least squares estimator
(equation \ref{eq:beta2sls}).
EViews uses the standard 3SLS formula (\ref{eq:cov3sls}) to
calculate the variance-covariance matrix of the 3SLS estimator.

>>>>

The Ordinary Least Squares (OLS) estimator of the system 
is obtained by
\begin{equation}
   \widehat{\beta} = \left( X'X \right)^{-1} X'y
\end{equation}
These estimates are efficient only if the disturbance terms are not
contemporaneously correlated, which means
$\sigma_{ij} = 0 \; \forall \; i \neq j$.
If the whole system is treated as one single equation,
the variance-covariance matrix of the estimated parameters is
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \sigma^2 \left( X'X \right)^{-1}
\end{equation}
with $\sigma^2 = E \left( u' u \right)$.
This assumes that the disturbances of all equations have the
same variance.

@@

The Weighted Least Squares (WLS) estimator of the system 
is obtained by
\begin{equation}
   \widehat{\beta} = \left( X' \Omega^{-1} X \right)^{-1} X' \Omega^{-1} y
\end{equation}
with $\Omega = \Sigma \otimes I$, 
$\sigma_{ij} = 0 \; \forall \; i \neq j$ and
$\sigma_{ii} = E \left( u_i' u_i \right)$.
Like the OLS estimates these estimates are only efficient
if the disturbance terms are not contemporaneously correlated.
The variance-covariance matrix of the estimated parameters is
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \left( X' \Omega^{-1} X \right)^{-1}
\end{equation}
If no cross-equation parameter restrictions are imposed,
the parameter estimates are equal to the OLS estimates.

@@

When the disturbances are contemporaneously correlated, a Generalized 
Least Squares (GLS) estimation leads to efficient parameter estimates.
In this case the GLS is generally called °Seemingly Unrelated Regression°
(SUR) (see \citealp{zellner62}). 
Note, while an unbiased OLS or WLS estimation requires only that 
the regressors and the disturbance terms of each single 
equation are uncorrelated
$( E \left[ u_i | X_i \right] = 0 \; \forall \; i )$,
a consistent SUR estimation requires that all disturbance terms and all 
regressors are uncorrelated
$( E \left[ u | X \right] = 0 )$.

The SUR estimator can be obtained by:%
\footnote{
To calculate $\Omega^{-1}$ it is not necessary to invert 
the (huge) $\Omega$ matrix, 
because it is sufficient to invert the (small) $\Sigma$ matrix and
calculate it by $\Omega^{-1} = \Sigma^{-1} \otimes I$
\citep[p.\ 342]{greene03}.
}
\begin{equation}
   \widehat{\beta} = \left( X' \Omega^{-1} X \right)^{-1} X' \Omega^{-1} y
\end{equation}
with $\Omega = \Sigma \otimes I$ and
$\sigma_{ij} = E \left( u_i' u_j \right)$.
And the variance-covariance matrix of the estimated parameters is
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \left( X' \Omega^{-1} X \right)^{-1}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Two-Stage least squares (2SLS)}

If the regressors of one or more equations are correlated 
with the disturbances ($E \left( u_i | X_i \right) \neq 0$), 
the estimated coefficients are biased.
This can be circumvented by a two-stage least squares (2SLS) estimation.
The instrumental variables for each equation $H_i$ 
can be either different or identical for all equations.
The instrumental variables of each equation may not be correlated with 
the disturbance terms of the corresponding equation 
($E \left( u_i | H_i \right) = 0$).

At the first stage new ('fitted') regressors are obtained by
\begin{equation}
   \widehat{X_i} = H_i \left( H_i' H_i \right)^{-1} H_i' X
\end{equation}
At the second stage the unbiased two-stage least squares estimates
of $\beta$ are obtained by:
\begin{equation}
   \widehat{\beta} = \left( \widehat{X}' \widehat{X} \right)^{-1} 
   \widehat{X}' y 
   \label{eq:beta2sls}
\end{equation}
If the whole system is treated as one single equation,
the variance-covariance matrix of the estimated parameters is
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \sigma^2 \left( \widehat{X}'
   \widehat{X} \right)^{-1}
\end{equation}
with $\sigma^2 = E \left( u' u \right)$.
If the disturbance terms of the individual equations
are allowed to have different variances, 
the variance-covariance matrix of the estimated parameters is
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \left( \widehat{X}' \Omega^{-1} 
   \widehat{X} \right)^{-1}
\end{equation}
with $\Omega = \Sigma \otimes I$, 
$\sigma_{ij} = 0 \; \forall \; i \neq j$ and
$\sigma_{ii} = E \left( u_i' u_i \right)$.

@@

The Weighted Two-Stage Least Squares (W2SLS) estimator of the system 
is obtained by
\begin{equation}
   \widehat{\beta} = \left( \widehat{X}' \Omega^{-1} \widehat{X} 
   \right)^{-1} \widehat{X}' \Omega^{-1} y
\end{equation}
with $\Omega = \Sigma \otimes I$, 
$\sigma_{ij} = 0 \; \forall \; i \neq j$ and
$\sigma_{ii} = E \left( u_i' u_i \right)$.
The variance-covariance matrix of the estimated parameters is
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \left( \widehat{X}' \Omega^{-1} 
   \widehat{X} \right)^{-1}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Three-Stage least squares (3SLS)}

If the regressors are correlated with the disturbances 
($E \left( u | X \right) \neq 0$) and 
the disturbances are contemporaneously correlated, 
a Generalized Least Squares (GLS) version of the two-stage least squares
estimation leads to consistent and efficient estimates.
This estimation procedure is generally called °Three-stage Least
Squares° (3SLS).

The standard 3SLS estimator can be obtained by:
\begin{equation}
   \widehat{\beta} = \left( \widehat{X}' \Omega^{-1} \widehat{X} 
   \right)^{-1} \widehat{X}' \Omega^{-1} y
\end{equation}
with $\Omega = \Sigma \otimes I$ and
$\sigma_{ij} = E \left( u_i' u_j \right)$.
Its variance-covariance matrix is:
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \left( \widehat{X}' \Omega^{-1} 
   \widehat{X} \right)^{-1}
   \label{eq:cov3sls}
\end{equation}
While an unbiased 2SLS or W2SLS estimation requires only that
the instrumental variables and the disturbance terms of each single 
equation are uncorrelated
$( E \left[ u_i | H_i \right]) = 0 \; \forall \; i )$,
\cite{schmidt90} points out that this estimator is only consistent 
if all disturbance terms and all instrumental variables are uncorrelated
$( E \left[ u | H \right]) = 0 )$
with
\begin{equation}
   H =
   \left[ \begin{array}{cccc}
      H_1 & 0 & \cdots & 0\\
      0 & H_2 & \cdots & 0\\
      \vdots & \vdots & \ddots & \vdots\\
      0 & 0 & \cdots & H_G
   \end{array}\right]
\end{equation}
Since there might be occasions where this cannot be avoided,
\cite{schmidt90} analyses other approaches to obtain 3SLS estimators:

One of these approaches is based on instrumental variable estimation
(IV-3SLS):
\begin{equation}
   \widehat{\beta} = \left( \widehat{X}' \Omega^{-1} X 
   \right)^{-1} \widehat{X}' \Omega^{-1} y
\end{equation}
The variance-covariance matrix of this IV-3SLS estimator is:
\begin{equation}
   Var \left[ \widehat{\beta} \right] = \left( \widehat{X}' \Omega^{-1} 
   X \right)^{-1}
\end{equation}
Another approach is based on the Generalized Method of Moments (GMM)
estimator (GMM-3SLS):
\begin{equation}
   \widehat{\beta} = \left( X' H \left( H' \Omega H \right)^{-1}
   H' X \right)^{-1} X' H \left( H' \Omega H \right)^{-1} H' y
   \label{eq:beta3slsGMM}
\end{equation}
The variance-covariance matrix of the GMM-3SLS estimator is:
\begin{equation}
   Var \left[ \widehat{\beta} \right] = 
   \left( X' H \left( H' \Omega H \right)^{-1} H' X \right)^{-1}
\end{equation}
A fourth approach developed by \cite{schmidt90} himself is:
\begin{equation}
   \widehat{\beta} = \left( \widehat{X}' \Omega^{-1} \widehat{X} 
   \right)^{-1} \widehat{X}' \Omega^{-1} 
   H \left( H' H \right)^{-1} H' y 
   \label{eq:beta3slsSchmidt}
\end{equation}
The variance-covariance matrix of this estimator is:
\begin{equation}
   Var \left[ \widehat{\beta} \right] = 
   \left( \widehat{X}' \Omega^{-1}  \widehat{X} \right)^{-1} 
   \widehat{X}' \Omega^{-1} H \left( H' H \right)^{-1} H' \Omega 
   H \left( H' H \right)^{-1} H' \Omega^{-1} \widehat{X}
   \left( \widehat{X}' \Omega^{-1}  \widehat{X} \right)^{-1}
\end{equation}
The econometrics software EViews uses following approach:
\begin{equation}
   \widehat{\beta} = \widehat{\beta}_{2SLS} + 
   \left( \widehat{X}' \Omega^{-1} \widehat{X} \right)^{-1} 
   \widehat{X}' \Omega^{-1} \left( y - X \widehat{\beta}_{2SLS} \right)
\end{equation}
with $\widehat{\beta}_{2SLS}$ is the two-stage least squares estimator
(equation \ref{eq:beta2sls}).
EViews uses the standard 3SLS formula (\ref{eq:cov3sls}) to
calculate the variance-covariance matrix of the 3SLS estimator.

### 2006-02-15 paper/trunk/systemfit_intro.tex
changed a few line breaks that each sentence starts in a new line now
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@591 591db2bd-e0a0-4d22-8c19-2192dbb7da31
several minor revisions in sections Introduction, Estimation, Restrictions, and Other Issues
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@593 591db2bd-e0a0-4d22-8c19-2192dbb7da31

Many theoretical models consist of more than one equation. These
models can contain variables that appear on the left-hand side in one
equation and on the right-hand side of another equation
(simultaneity-bias). The system can have non-zero off-diagonal
covariance elements resulting from related disturbnaces
(contemporaneous correlation). Ignoring contemporaneous correlation
leads to inefficient parameter estimates \citep{zellner62} and
ignoring simultaneity-bias can lead to inconsistent parameter
estimates.

@@

Another reason to estimate an equation system simultaneously are
cross-equation parameter restrictions. These restrictions can be
tested and/or imposed only in a simultaneous estimation approach
especially the economic theory suggests many cross-equation
restrictions.

@@

% For all of the methods developed in the package, the disturbnaces of
% the individual equations are assumed to be independent and identially
% distributed (iid).  
% In the future, we would like to add the ability to fit equations were
% the disturbances are serially correlated (wikins 1969).

The \pkg{systemfit} package provides the capability to estimate
linear and non-linear equation systems in \proglang{R}.
Although linear and non-linear equation systems can be estimated
with several other statistical and econometric software packages
(e.g.\ \proglang{SAS}, \proglang{EViews}, \proglang{TSP}),
\pkg{systemfit} has several advantages.
First, all estimation procedures are publicly available in the source code.
Second, the estimation algorithms can be easily modified to meet specific
requirements.
Third, the (advanced) user can control estimation details generally
not available in other software packages by overriding reasonable defaults.

>>>>

Many theoretical models consist of more than one equation.
These models can contain variables that appear on the left-hand side in one
equation and on the right-hand side of another equation (simultaneity-bias).
The system can have non-zero off-diagonal covariance elements resulting from
related disturbances (contemporaneous correlation).
Ignoring contemporaneous correlation leads to inefficient parameter
estimates \citep{zellner62} and ignoring simultaneity-bias can lead to
inconsistent parameter estimates.

@@

Another reason to estimate an equation system simultaneously are
cross-equation parameter restrictions.
These restrictions can be tested and/or imposed only in a simultaneous
estimation approach.
Especially the economic theory suggests many cross-equation parameter
restrictions (e.g.\ the symmetry restriction in demand models).

@@

% For all of the methods developed in the package, the disturbances of
% the individual equations are assumed to be independent and identially
% distributed (iid).  
% In the future, we would like to add the ability to fit equations were
% the disturbances are serially correlated (wikins 1969).

The \pkg{systemfit} package provides the capability to estimate
linear and non-linear equation systems in \proglang{R}
\citep{r-project}.
Although linear and non-linear equation systems can be estimated
with several other statistical and econometric software packages
(e.g.\ \proglang{SAS}, \proglang{EViews}, \proglang{TSP}),
\pkg{systemfit} has several advantages.
First, all estimation procedures are publicly available in the source code.
Second, the estimation algorithms can be easily modified to meet specific
requirements.
Third, the (advanced) user can control estimation details generally
not available in other software packages by overriding reasonable defaults.

### 2006-02-15 paper/trunk/systemfit_other.tex
several minor revisions in sections Introduction, Estimation, Restrictions, and Other Issues
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@593 591db2bd-e0a0-4d22-8c19-2192dbb7da31

Since the true residuals of the estimated equations are generally not known,
the true covariance matrix of the residuals cannot be determined.
Thus, this covariance matrix must be calculated from the
\emph{estimated} residuals.

The estimated covariance matrix of the residuals ($\widehat{\Sigma} = 
\left[ \widehat{\sigma}_{ij} \right]$) 
is often calculated by
\begin{equation}
   \widehat{\sigma}_{ij} = \frac{ \widehat{u}_i' \widehat{u}_j }{ T }
   \label{eq:rcov0}
\end{equation}
where $T$ is the number of observations in each equation.

However, this estimator is biased, because it is not corrected for
degrees of freedom.
The usual single-equation procedure to correct for degrees of freedom
cannot always be applied, because the number of regressors in each equation
might differ.
Two alternative approaches to calculate the residual variance-covariance
matrix are
\begin{equation}
   \widehat{\sigma}_{ij} = \frac{ \widehat{u}_i' \widehat{u}_j }
   { \sqrt{ \left( T - K_i \right) \cdot \left( T - K_j \right) } }
   \label{eq:rcov1}
\end{equation}
and
\begin{equation}
   \widehat{\sigma}_{ij} = \frac{ \widehat{u}_i' \widehat{u}_j }
   { T - \max \left( K_i , K_j \right) }
\end{equation}
where $K_i$ and $K_j$ are the number of regressors in equation
$i$ and $j$, respectively.

However, this yields an unbiased estimator only if $K_i = K_j$ 
\citep[p.\ 469]{judge85}. 
% Greene (2002, p. 344) says that the second is unbiased if i = j or K_i = K_j,
% whereas the first is unbiased only if i = j. 
% However, if K_i = K_j the first and the second are equal.
% Why is the first biased if K_i = K_j ???????????

@@

The WLS, SUR, W2SLS and 3SLS parameter estimates are consistent,
if the estimated residual variance-covariance matrix is calculated
using the residuals from a first step OLS or 2SLS estimation.
It is also possible to iterate the estimation procedure and 
calculate the residual variance-covariance matrix from the
residuals of the previous iteration. 
If equation (\ref{eq:rcov0}) is applied to calculate the estimated
residual variance-covariance matrix, the results are maximum
likelihood \citep[p.\ 345]{greene02}.

@@

The goodness of fit of the whole system can be measured by the
McElroy's $R^2$ value \citep{mcelroy77}: 
% also: \citep[p.\ 345]{greene02}
\begin{equation}
   R_*^2 = 1 - \frac{ \widehat{u}' \widehat{ \Omega }^{-1} \widehat{u} }
   { y' \left( \widehat{ \Sigma }^{-1} \otimes
   \left( I - \frac{i i'}{T} \right) \right) y }
\end{equation}
where $T$ is the number of observations in each equation,
$I$ is an $T \times T$ identity matrix and 
$i$ is a column vector of $T$ ones.

>>>>

Since the true residuals of the estimated equations are generally not known,
the true covariance matrix of the residuals cannot be determined.
Thus, this covariance matrix must be calculated from the
\emph{estimated} residuals. 
Generally, the estimated covariance matrix of the residuals
($\widehat{\Sigma} = \left[ \widehat{\sigma}_{ij} \right]$)
can be calculated from the residuals of a first-step OLS or 2SLS estimation.
The following formula is often applied:
\begin{equation}
   \widehat{\sigma}_{ij} = \frac{ \widehat{u}_i' \widehat{u}_j }{ T }
   \label{eq:rcov0}
\end{equation}
where $T$ is the number of observations in each equation.

However, this estimator is biased, because it is not corrected for
degrees of freedom.
The usual single-equation procedure to correct for degrees of freedom
cannot always be applied, because the number of regressors in each equation
might differ.
Two alternative approaches to calculate the residual variance-covariance
matrix are
\begin{equation}
   \widehat{\sigma}_{ij} = \frac{ \widehat{u}_i' \widehat{u}_j }
   { \sqrt{ \left( T - K_i \right) \cdot \left( T - K_j \right) } }
   \label{eq:rcov1}
\end{equation}
and
\begin{equation}
   \widehat{\sigma}_{ij} = \frac{ \widehat{u}_i' \widehat{u}_j }
   { T - \max \left( K_i , K_j \right) }
   \label{eq:rcov3}
\end{equation}
where $K_i$ and $K_j$ are the number of regressors in equation
$i$ and $j$, respectively.

However, this yields an unbiased estimator only if $K_i = K_j$ 
\citep[p.\ 469]{judge85}. 
% Greene (2003, p. 344) says that the second is unbiased if i = j or K_i = K_j,
% whereas the first is unbiased only if i = j. 
% However, if K_i = K_j the first and the second are equal.
% Why is the first biased if K_i = K_j ???????????

@@

The WLS, SUR, W2SLS and 3SLS parameter estimates are consistent,
if the estimated residual variance-covariance matrix is calculated
using the residuals from a first-step OLS or 2SLS estimation.
It is also possible to iterate the estimation procedure and 
calculate the residual variance-covariance matrix from the
residuals of the previous iteration. 
If equation (\ref{eq:rcov0}) is applied to calculate the estimated
residual variance-covariance matrix, the results are maximum
likelihood \citep[p.\ 345]{greene03}.

@@

The goodness of fit of the whole system can be measured by the
McElroy's $R^2$ value \citep{mcelroy77}: 
% also: \citep[p.\ 345]{greene03}
\begin{equation}
   R_*^2 = 1 - \frac{ \widehat{u}' \widehat{ \Omega }^{-1} \widehat{u} }
   { y' \left( \widehat{ \Sigma }^{-1} \otimes
   \left( I - \frac{i i'}{T} \right) \right) y }
\end{equation}
where $T$ is the number of observations in each equation,
$I$ is an $T \times T$ identity matrix and 
$i$ is a column vector of $T$ ones.
